---
title: "p8105_hw3_bz2570"
author: "Boran Zhai"
date: "2025-10-09"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(patchwork)
```

## Problem 1
#### Import the data
```{r}
library(p8105.datasets)
data("instacart")
```

#### A short description of the dataset (Inline R code)
The dataset instacart contains `r nrow(instacart)` rows and `r ncol(instacart)` columns, with the following variables: `r paste(names(instacart), collapse = ", ")`. <br/> 

The dataset has the following structure:<br/>
```{r}
str(instacart)
```

Description for some key variables:<br/>
order_id: The unique identification of order for every user <br/>
product_id: The unique identification for every product <br/>
user_id: The unique identification of every user <br/>
reordered: The number of reordered product <br/>
aisle_id: The identification of aisle <br/>
aisle: Product subcategory <br/>
department: Major product category <br/>
add_to_cart_order: Sequence position when item was added to cart <br/>
days_since_prior_order: Days since user's previous order <br/>

Some illustrative examples of observations are as follows:
```{r}
instacart |> 
head() |> 
  knitr::kable()
```
For the product Bulgarian Yogurt, the data shows that it was the first item added to the cart by user 112108 in their 4th overall order. This purchase was a reorder, indicating the user's prior satisfaction with the product. The order was placed on a Thursday at 10 AM, coming 9 days after their previous order. This yogurt is categorized in the yogurt aisle within the dairy eggs department.

#### 1. How many aisles, and which are most popular?
```{r}
aisles <- instacart |> 
  group_by(aisle) |> 
  summarize(total_items = n()) |> 
  arrange(desc(total_items))

aisles |> 
  head(5) |> 
  knitr::kable(digits = 1)
```
There are `r nrow(aisles)` aisles. The top 5 aisles are `r paste(head(aisles$aisle, 5), collapse = ", ")`. <br/>

#### 2. Make a plot that shows the number of items ordered in each aisle, limiting this to aisles with more than 10000 items ordered. 
```{r}
aisle_data <- instacart |> 
  group_by(aisle) |> 
  summarize(total_items = n()) |> 
  filter(total_items > 10000) |> 
  mutate(aisle = fct_reorder(aisle, total_items))

aisle_plot <- aisle_data |> 
  ggplot(aes(x = total_items, y = aisle, fill = total_items)) +
  geom_col(alpha = 0.8) +
  scale_fill_gradient(low = "#8264CC", high = "#4a1486") +
  labs(
    title = "Number of Items Ordered in Each Aisle",
    subtitle = "(Aisles with more than 10000 items ordered)",
    x = "number of items ordered",
    y = "aisle",
  ) +
  theme_minimal()
print(aisle_plot)
```
<br/>The plot shows the number of items ordered in each aisle, and it is ordered by descending order.

#### 3. Make a table showing the three most popular items in each of the aisles “baking ingredients”, “dog food care”, and “packaged vegetables fruits”. Include the number of times each item is ordered in your table.
```{r}
popular_items_table <- instacart |> 
  filter(aisle %in% c("baking ingredients", "dog food care", "packaged vegetables fruits")) |> 
  count(aisle, product_name, name = "order_num") |> 
  group_by(aisle) |> 
  slice_max(order_num, n = 3) |> 
  arrange(aisle, desc(order_num))
popular_items_table |> 
  knitr::kable()
```
In the baking ingredients aisle, the most popular item is `r popular_items_table |> filter(aisle == "baking ingredients") |> slice(1) |> pull(product_name)` with `r popular_items_table |> filter(aisle == "baking ingredients") |> slice(1) |> pull(order_num)` orders. <br/>

In the dog food care aisle, the most popular item is `r popular_items_table |> filter(aisle == "dog food care") |> slice(1) |> pull(product_name)` with `r popular_items_table |> filter(aisle == "dog food care") |> slice(1) |> pull(order_num)` orders. <br/>

In the packaged vegetables fruits aisle, the most popular item is `r popular_items_table |> filter(aisle == "packaged vegetables fruits") |> slice(1) |> pull(product_name)` with `r popular_items_table |> filter(aisle == "packaged vegetables fruits") |> slice(1) |> pull(order_num)` orders. <br/>

#### 4. Make a table showing the mean hour of the day at which Pink Lady Apples and Coffee Ice Cream are ordered on each day of the week; format this table for human readers (i.e. produce a 2 x 7 table).
```{r}
mean_order_hod <- instacart |>       # mean hour of the day
  filter(product_name %in% c("Pink Lady Apples", "Coffee Ice Cream")) |> 
  group_by(product_name, order_dow) |> 
  summarize(mean_order_hour = mean(order_hour_of_day, na.rm = TRUE), .groups = "drop") |> 
  mutate(mean_order_hour = round(mean_order_hour, 2)) |> 
  pivot_wider(
    names_from = order_dow,
    values_from = mean_order_hour
  )
names(mean_order_hod) <- c("Product", "Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday")
mean_order_hod |> 
  knitr::kable(caption = "The mean hour of the day at which Pink Lady Apples and Coffee Ice Cream are ordered on each day of the week")
```
The analysis shows that Pink Lady Apples are typically ordered around `r round(mean(as.numeric(mean_order_hod[2,2:8])), 0)`:00, while Coffee Ice Cream is ordered around `r round(mean(as.numeric(mean_order_hod[1,2:8])), 0)`:00 on average, indicating that pink lady apples are always purchased earlier in the day than coffee ice-cream.

## Problem 2

#### Import, clean, and tidy Zillow datasets
Import and clean ZIP code dataset
```{r}
zip_codes_tidy <- read_csv("./zillow_data/Zip Codes.csv",
                           na = c(".", "NA", ""))|>
  janitor::clean_names() |>   
  mutate(zip_code = as.character(zip_code),
         borough = county
         ) |>
  select(-county) |> 
  distinct(zip_code, .keep_all = TRUE) |>
  relocate(zip_code) |> 
  arrange(zip_code) 
```

Import and clean ZIP ZORI dataset
```{r}
zip_zori_tidy <- read_csv("./zillow_data/Zip_zori_uc_sfrcondomfr_sm_month_NYC.csv",
                     na = c(".", "NA", "")) |>
  janitor::clean_names() |> 
  pivot_longer(
  cols = starts_with("x"),
  names_to = "date",        
  values_to = "zori"        
  ) |>
  mutate(
    date = str_remove(date, "^x"),   
    date = str_replace_all(date, "_", "-"),  
    date = as.Date(date),               
    year = year(date),
    county_name = str_remove(county_name, " County$")
    ) |>  
  filter(date >= as.Date("2015-01-01"), date <= as.Date("2024-08-31")) |>
  relocate(region_name) |> 
  arrange(region_name) |>  
  filter(!is.na(zori))
```

Merge the datasets
```{r}
resulting_tidy_data <- zip_zori_tidy |>
  mutate(region_name = as.character(region_name)) |> 
  left_join(zip_codes_tidy, by = c("region_name" = "zip_code")) |>
  filter(!is.na(borough))
```

#### There are 116 months between January 2015 and August 2024.<br/>
```{r}
zip_obs_count <- zip_zori_tidy  |> 
  count(region_name, name = "observation_count") |> 
  arrange(observation_count)
# How many ZIP codes are observed 116 times? 
zip_obs_116 <- zip_obs_count |> 
  filter(observation_count == 116) |> 
  nrow() 
# How many are observed fewer than 10 times? 
zip_obs_fewer_10 <- zip_obs_count  |> 
  filter(observation_count < 10) |> 
  nrow() 
```
The data shows that `r zip_obs_116` ZIP codes are observed 116 times from January 2015 to August 2024, while `r zip_obs_fewer_10` ZIP codes have fewer than 10 observations. This disparity can be attributed to fundamental differences in neighborhood characteristics and data collection practices. Established residential areas with stable rental markets and consistent transaction volumes support regular data reporting, whereas transitional neighborhoods, developing areas, or regions with limited rental activity produce intermittent data due to lower transaction frequency, evolving housing stock, or challenges in systematic data capture.

#### Create a rtable showing the average rental price in each borough and year
```{r}
# create a table
rental_price_table <- resulting_tidy_data |>
  group_by(borough, year) |>
  summarize(avg_rent = mean(zori, na.rm = TRUE), .groups = "drop")|>
  pivot_wider(
    names_from = year,
    values_from = avg_rent
  ) |>
  knitr::kable(caption = "Average rental price in each borough and year", digits = 0)
rental_price_table

# comment the trends
trend <- resulting_tidy_data |>
  group_by(borough, year) |>
  summarize(avg_zori = mean(zori, na.rm = TRUE), .groups = "drop") |>
  group_by(borough) |>
  summarize(
    start_year = min(year),
    end_year = max(year),
    start_price = first(avg_zori[order(year)]),
    end_price = last(avg_zori[order(year)]),
    pct_change = (end_price - start_price) / start_price * 100,
    years_covered = n_distinct(year),
    .groups = "drop"
  ) |> 
  arrange(desc(pct_change))
```
The rental price analysis reveals distinct market dynamics across NYC's `r n_distinct(resulting_tidy_data$borough)` boroughs from `r min(trend$start_year)` to `r max(trend$end_year)`. `r trend$borough[1]` showed the strongest growth with a `r round(trend$pct_change[1], 1)`% increase in average rental price, rising from $`r round(trend$start_price[1], 0)` to $`r round(trend$end_price[1], 0)`. `r trend$borough[2]` experienced a `r round(trend$pct_change[2], 1)`% increase, with prices moving from $`r round(trend$start_price[2], 0)` to $`r round(trend$end_price[2], 0)`. `r trend$borough[3]` saw a `r round(trend$pct_change[3], 1)`% rise, from $`r round(trend$start_price[3], 0)` to $`r round(trend$end_price[3], 0)`. `r trend$borough[4]` showed a `r round(trend$pct_change[4], 1)`% growth, increasing from $`r round(trend$start_price[4], 0)` to $`r round(trend$end_price[4], 0)`. In contrast, `r trend$borough[nrow(trend)]` experienced more moderate changes at `r round(trend$pct_change[nrow(trend)], 1)`% growth, with average rental price moving from $`r round(trend$start_price[nrow(trend)], 0)` to $`r round(trend$end_price[nrow(trend)], 0)`.

#### Make a plot showing NYC Rental Prices within ZIP codes for all available years.
```{r}
# Create rental price plot using techniques from your weather data examples
rental_price_plot <- resulting_tidy_data |> 
  ggplot(aes(x = date, y = zori, color = borough)) +
  geom_point(alpha = 0.1) +
  geom_smooth(se = FALSE, size = 1) +
  facet_wrap(. ~ borough) +
  labs(
    title = "NYC Rental Prices by Borough (2015-2024)",
    x = "Year",
    y = "Rental Price",
    color = "Borough",
    caption = "Data from the Zillow ZORI Dataset"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    plot.margin = margin(20, 20, 20, 20)
  ) +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  scale_color_viridis_d()
print(rental_price_plot)
```

Comment:<br/>
A plot displaying New York City rental prices by borough from 2015 to 2024 reveals distinct trends across different boroughs. The Bronx (purple) exhibits a steady upward trajectory in rental prices throughout the period. Kings County (blue) shows a pattern of gradual increase, with a more pronounced rise in the latter years. New York County (teal) demonstrates notable fluctuations yet sustains a long-term upward trend, and its overall rental price levels are relatively high. Queens (green) features a consistent upward trend, with rental prices increasing steadily from a moderate initial baseline. Richmond County (yellow) has missing data in the early years; when data become available, it starts with relatively low rental prices and undergoes a gradual increase in the later years, although its overall rental level remains lower compared to most other boroughs.

#### Compute and make a plot showingthe average rental price within each ZIP code over each month in 2023
```{r}
# Compute average rental price for each ZIP code in 2023
rental_prize_2023 <- resulting_tidy_data |> 
  filter(year(date) == 2023) |> 
  group_by(borough, zip_code = region_name) |> 
  summarize(
    avg_rent_2023 = mean(zori, na.rm = TRUE),
    .groups = "drop"
  )

# Create distribution plot using boxplots
rental_boxplot_2023 <- rental_prize_2023 |> 
  ggplot(aes(x = borough, y = avg_rent_2023)) +
  geom_boxplot(aes(fill = borough), alpha = 0.7) +
  labs(
    title = "Distribution of Average Rental Prices by Borough (2023)",
    x = "Borough", 
    y = "Average Rental Price",
    caption = "Data from Zillow Dataset"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16)
  ) +
  scale_fill_manual(values = c("#BDB2FF","#FFADAD", "#FDFFB6", "#CAFFBF","#9BF6FF"))

print(rental_boxplot_2023)
```
Comment: <br/>
This box plot shows the distribution of average rental prices across NYC boroughs in 2023. New York Borough has the highest average rental prices, featuring the widest range with notable high outliers and a median over $4000. Kings Borough has a lower median and range than New York but higher ones than the Bronx, Queens and Richmond. The Bronx has a median less than 2500. Queens has a median slightly above that of the Bronx, with a narrower range. Richmond Borough has the lowest average rental prices and the narrowest distribution among all boroughs.

#### Combine and export the plots
```{r}
combined_rental_plot <- rental_price_plot / rental_boxplot_2023
print(combined_rental_plot)
ggsave("results/combined_rental_plot.png", combined_rental_plot, width = 15, height = 20)
```

## Problem 3
#### Load, tidy, merge, and otherwise organize the data sets. 
```{r}
# Load and tidy the data sets
accel__clean = read_csv("./data/nhanes_accel.csv") |>
  janitor::clean_names()

covar_clean <- read_csv("./data/nhanes_covar.csv", skip = 4) |>
  janitor::clean_names() |>
  filter(age >= 21) |>
  drop_na() |>
  mutate(
    sex = factor(sex, levels = c(1, 2), labels = c("Male", "Female")),
    education = factor(education, levels = 1:3, 
                      labels = c("Less than high school", 
                                "High school equivalent", 
                                "More than high school"))
  )
# Merge the data sets
nhanes_clean <- covar_clean |>
  left_join(accel__clean, by = c("seqn"))
```

#### Create a table and a visualization
```{r}
# Produce a table for the number of men and women in each education category
edu_sex_table <- nhanes_clean |>
  count(education, sex) |>
  pivot_wider(names_from = sex, values_from = n)
knitr::kable(edu_sex_table,caption = "The table for the number of men and women in each education category")

# Create a visualization of the age distributions for men and women in each education category
age_dist_plot <- nhanes_clean |>
  ggplot(aes(x = age, fill = sex)) +
  geom_histogram(position = "dodge", alpha = 0.8, binwidth = 4) +
  facet_grid(education ~ .) +
  labs(
    title = "Age Distribution by Sex and Education Level",
    x = "Age",
    y = "Count",
    fill = "Sex"
  )+
  theme_minimal() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16)
  ) 
print(age_dist_plot)
```
Comment:<br/>
The table indicates that for “Less than high school”, male (27) and female (28) counts are nearly equal; “High school equivalent” has more males (35) than females (23); and “More than high school” has slightly more females (59) than males (56). The histogram visualizes age distributions by sex across education levels: “Less than high school” shows relatively balanced sex counts across ages; “High school equivalent” has higher male counts in some age bins; and “More than high school” features a notable peak in female counts for younger ages (around 30–40) with females often matching or exceeding male counts in most age groups.

#### Aggregate across minutes to create a total activity variable for each participant
```{r}
nhanes_total_activity <- nhanes_clean |>
  mutate(total_activity = rowSums(across(starts_with("min")), na.rm = TRUE)) |>
  select(seqn, sex, age, bmi, education, total_activity)
```

#### Plot total activities (y-axis) against age (x-axis)
```{r}
total_activity_plot <- nhanes_total_activity |>
  ggplot(aes(x = age, y = total_activity, color = sex)) + 
  geom_point(alpha = 0.5) + 
  geom_smooth(se = FALSE) + 
  facet_wrap(~ education) + 
  labs(
    title = "Total activities against age", 
    x = "Age", 
    y = "Total Activity (MIMS)", 
    color = "Sex"
  ) +
  theme_minimal()+
  theme(
    legend.position = "bottom",
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14)
  )
print(total_activity_plot)
```
Comment: <br/>
This plot shows total daily activity against age, comparing males (represented by the red line and points) and females (represented by the teal line and points) across three education groups. For individuals with less than a high school education, total activity decreases with age for both sexes, where males start with higher activity levels but see a sharper decline after 60. Among those with a high school equivalent education, activity peaks around middle age (40) for both male and female, and females maintain higher activity levels. For people with more than a high school education, activity remains more stable across different ages untile 60, and females consistently have higher total activity than males.

#### Create three-panel plot for 24-hour activity time courses for each education level and use color to indicate sex
```{r}
activity_24_plot <- nhanes_clean |>
  pivot_longer(
    cols = starts_with("min"), 
    names_to = "minute", 
    values_to = "mims_value",
    names_prefix = "min"
  ) |>
  mutate(minute = as.numeric(minute)) |> 
  group_by(education, sex, minute) |>
  summarize(mean_mims = mean(mims_value, na.rm = TRUE), 
            .groups = "drop") |>
  ggplot(aes(x = minute, y = mean_mims, color = sex)) + 
  geom_line(alpha = 0.7) + 
  geom_smooth(se = FALSE, linewidth = 1.2) +
  facet_wrap(~ education, nrow = 1) +
  labs(
    title = "24-Hour Activity Time Courses by Education Level and Sex", 
    x = "Time of Day", 
    y = "Average MIMS", 
    color = "Sex"
  ) + 
  scale_x_continuous(
    breaks = c(0, 360, 720, 1080, 1440), 
    labels = c("0:00", "6:00", "12:00", "18:00", "24:00") 
  ) +
    theme_minimal()+
    theme(
      legend.position = "bottom",
      plot.title = element_text(hjust = 0.5, face = "bold", size = 14)
    )
print(activity_24_plot)
```
Comment: <br/>
This plot shows 24-hour activity time courses across three education levels, with color distinguishing sex (red for males, teal for females). Across all education levels, both sexes follow a similar daily pattern that activity rises in the morning, peaks around midday, and declines in the evening. However, females consistently show higher average activity than males during peak hours (around midday) in every education group. Additionally, the magnitude of peak activity varies by education, with the “Less than high school” group having the highest peak. Overall, while the daily activity rhythm is similar, gender differences and the changes in peak activity amplitude vary with education level.

```{r}
ggsave("results/age_dist_plot.png", age_dist_plot)
ggsave("results/total_activity_plot.png", total_activity_plot)
ggsave("results/activity_24_plot.png", activity_24_plot)
```

